Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-25 tex/disscussion.tex
Rewording: All sections finished but 5.

\section{Discussion and conclusion}\label{sec:disscussion}

Traditionally, the proof of the classical $H$-theorem involves assuming that the
distribution function is spatially homogeneous, and the proof starts with
considering particular systems that are out of equilibrium, but in such a manner
that their distribution functions are still homogeneous. This out-of-equilibrium systems
are conceived considering the micro-states of the system, and the $H$-theorem determines
the microscopic evolution of the system towards a state with the most-probable distribution function,
given a thermodynamic state. 
This approach does not cover systems with spatial inhomogeneities.
To address this issue, we
proposed a framework that may be useful to describe the time-evolution of systems
initially non-homogeneous. For this, we divided the
system into small cells, so as to conceive a system wherein we accept the local equilibrium
hypothesis to be valid for each cell, but in such a manner that the global system is not homogeneous. Systems
that satisfy the previous conditions will evolve towards global equilibrium, and the evolution is
determined by the relations $d\mathcal{H}'/dt\leq0$, Eq.~(\ref{eq:cHdef}),
and $d\mathcal{H}/dt\leq0$, Eq.~(\ref{eq:qHdef}), for classical and quantum gases, respectively.
In some aspects, this can be considered as an extension or as a potential equivalent of the $H$ theorem
for systems out-of-equilibrium in the spatial sense, \textit{i.e.} systems whose thermodynamic
variables are not spatially homogeneous.

The classical and quantum $H$-functionals, $\mathcal{H}'$ and $\mathcal{H}$, respectively,
correctly recover the most-probable distribution
functions both in out-of-equilibrium states (locally) and when the system attains the global
equilibrium state. This enables the use of all the machinery of thermodynamics, including the
proper definition of state variables as functions of time for each cell. In addition, our approach
does not require to know the specific mechanisms by which energy or particles are transferred from one
cell to its neighbors. Instead, the time-evolution from the out-of-equilibrium to equilibrium states are
governed by monotonic functions that account for deviations from the global equilibrium.

The specific forms of the monotonic functions $g'_M$ and $g_{nM}$ for classical and quantum systems will be the
subject of our future work. However, some of their properties can be foreseen, \textit{e.g.} they must
be consistent with the system relaxation times, they must be consistent with the mechanisms of 
energy transfer between open systems, they must be related to the cell thermodynamic variables, etc.

An important aspect of the framework proposed in this work is related to the entropy of systems
out-of-equlibrium. Because the functionals $\mathcal{H}'$ and $\mathcal{H}$ can be related to
the entropy of diluted gases, either classical or quantum, the fact that these functionals are defined
over a system divided into cells enables their use for defining the entropy
of other out-of-equilibrium systems. In particular, and 
derived from our previous work (\textit{e.g.} \cite{bib:nicolas2020,bib:nicolas2016}), the
$\mathcal H'$ and $\mathcal H$ functionals may serve to describe the entropy,
as well as the entropy generation, occurring during the
growth of complex physical systems, such as fractals. Possibly, studying these systems might also
shed light on the functional form of $g'_M$ and $g_{nM}$.

In conclusion, we have proposed a framework for studying classical and quantum out-of-equilibrium
diluted gases, whose initial
state is inhomogeneous in nature. This framework consists, briefly, of dividing the system into
a set of small cells, and assuming that each cell is in equilibrium. Subsequently,
$H$-functionals, $\mathcal H$, can be defined
such that the time evolution towards equilibrium resembles the Boltzmann $H$-theorem: $d\mathcal H/dt\leq0$,
and the equality condition is satisfied when the system attains the global equlibrium.
The distribution functions,
stemming from the $\mathcal H$ functionals, at any time, $t$, are the respective equilibrium distribution functions,
which can be Maxwell-Boltzmann, Fermi-Dirac or Bose-Einstein, depending on the classical or quantum nature
of the gas under consideration. Furthermore,
the distribution function of each cell evolves so as to match the global distribution function in equilibrium.
A proof of the relation $d\mathcal H/dt\leq0$ is performed through the variational method,
the functional form of the distribution
functions, in terms of a monotonic functions $g$, are defined, and some of the properties of $g$ are discussed.
The functionals $\mathcal H$ can be associated with the entropy of the system under consideration, hence we
refer to this entropy as the variational entropy of the system. Finally,
a potential application of this framework is outlined briefly.

>>>>

\section{Comments and remarks}\label{sec:disscussion}

The demonstration of the classical $H$-theorem usually starts by assuming that the gas,
despite being initially out of equilibrium, can be described by a
spatially homogeneous distribution function. 
Subsequently, the time-evolution of the system occurs in such a manner that $dH/dt \leq 0$. 
Therefore, this approach does not cover the evolution to equilibrium of systems with spatial inhomogeneities.
To address this issue, in this article we
proposed a framework that may be useful to describe the time-evolution of systems
initially non-homogeneous. To this end, we divided the
system into small cells, so as to conceive a system wherein the local equilibrium
hypothesis is valid in each cell, but in such a manner that the total system is not homogeneous. Systems
that satisfy the previous conditions will evolve towards equilibrium, and the evolution
occurs according to $d\mathcal{H}'/dt\leq0$, Eq.~(\ref{eq:cHdef}),
and $d\mathcal{H}/dt\leq0$, Eq.~(\ref{eq:qHdef}), for classical and quantum gases, respectively.
Consequently, this approach can be considered as an extension of the $H$-theorem
for more realistic out-of-equilibrium systems.

The classical and quantum $H$-functionals, $\mathcal{H}'$ and $\mathcal{H}$, respectively,
correctly recover the most-probable distribution
functions both in out-of-equilibrium states (locally) and when the system attains the global
equilibrium state. \commb{The relaxation process of the system is described by
monotonic functions that account for deviations from the global equilibrium.}

It is clear that for describing the relaxation process of a concrete system,
it is necessary to know, at least with some approximation level, the specific forms of the
monotonic functions $g'_M$ and $g_{nM}$, for classical and quantum systems.
Whereas the complete analysis of these functions is out of the scope of the present work,
some of their properties can be foreseen, \textit{e.g.} they must
be consistent both with the system relaxation times and with the mechanisms of 
energy transfer between cells.

An important aspect of the framework proposed in this work is related to the entropy of systems
out-of-equlibrium. Because the functionals $\mathcal{H}'$ and $\mathcal{H}$ can be related to
the entropy of diluted gases, either classical or quantum, the fact that these functionals are defined
over a system divided into cells enables their use for defining the entropy
of out-of-equilibrium systems, other than diluted gases. In particular, and 
derived from our previous work (\textit{e.g.} \cite{bib:nicolas2020,bib:nicolas2016}), the
$\mathcal H'$ and $\mathcal H$ functionals may serve to describe the entropy,
as well as the entropy generation, occurring during the
growth of complex physical systems, such as fractals. Possibly, studying these systems might also
shed light on the functional form of $g'_M$ and $g_{nM}$.

In summary, we proposed a variational procedure to demonstrate the classical and
quantum $H$-theorems, which made possible to describe, at a mesoscopic local view
(cell-scale), the time-evolution of an out-of-equilibrium and spatially
non-homogeneous system, towards the equilibrium condition. In principle, this
approach would render possible to investigate the transport phenomena, inherent to the
equilibration process, occurring in a system that started from an spatially inhomogeneous
out-of-equilibrium initial condition.

### 2021-01-25 tex/introduction.tex
Rewording: sec. 6 (disc. and concl).
Rewording: All sections finished but 5.

In the literature, the theoretical basis that allows us to describe equilibrium
systems in the classical scheme is well-established.
The procedures to describe
equilibrium systems are well-known and work in almost every system in nature
(Thermodynamics for a phenomenological description and statistical mechanics
for a microscopic description). In contrast, a unified mathematical model to
describe the evolution of any system to an equilibrium state has not developed.
Nevertheless, developments can be found in the literature and describe the
evolution to equilibrium to almost any system (the internal behavior of glasses
\cite{bib:zanotto2018} and the entropy measurement
\cite{bib:schmelzer2018,bib:nemilov2018} are some examples where the recent
developments do not work.) Examples of those developments can be the Onsager
formulation (called linear thermodynamics)
\cite{bib:keizer1987,bib:onsager1931}, and the Kinetic Theory of dilute gases
developed by Boltzmann.

In the Kinetic Theory of dilute gases, the behavior of a dilute classical gas
was described by Boltzmann through the Boltzmann transport equation.
Furthermore, the evolution of an equilibrium state for any system was proved by
means the Boltzmann's $H$-theorem (or classical $H$-theorem) and the
spatial-homogeneous distribution function hypothesis.
However, there are some aspects in the classical $H$-theorem and the
Boltzmann's $H$-functional that deserve some review to accomplish better
consistency, for example, the modification of the $H$-theorem in stochastic
trajectories, violations in the second law of thermodynamics, and the relation
between Shannon's measure of information and the Boltzmann's entropy %[6-10].
\cite{bib:nemilov2018,bib:keizer1987,bib:onsager1931,bib:brown2008,bib:dragoljub2009}.

To obtain the well-known correspondence principle between classical mechanics
and quantum mechanics, a quantum version of the Boltzmann's $H$-theorem (or
quantum $H$-theorem) and the quantum analogous to the Boltzmann transport
equation is necessary such that in the appropriate limit, the  classical $H$
theorem and the Boltzmann transport equation should be recovered.

One of the first developments in this field was developed by Tolman
\cite{bib:tolman}. He proposes one of the first quantum versions of the
Boltzmann's $H$-theorem using the probability transition relation, the random
phases hypothesis, and a homogeneous distribution function. The equation of
motion for the occupation number (necessary to prove the quantum $H$-theorem in
the same way that Boltzmann's procedure) was obtained by applying the time
perturbation theory. However, to obtain the flux relations in analogy to the
Boltzmann's method, a quantum version of the transport equation that includes a
non-homogeneous distribution function is required.

On the other hand, some works in the quantum operator formalism attempt to
describe quantum transport phenomena using the Hamiltonian of the system and
the master equation (analogous to the Boltzmann transport equation)
\cite{bib:gorban2014,bib:bennaim2017,bib:tolman,bib:li2019}. %[12-15]
Besides, some authors construct the $H$-functional and develop the proof of the
quantum $H$-theorem \cite{bib:silva2010,bib:deroeck2006,bib:grabert1974}.
However, the homogeneous distribution function hypothesis and the
correspondence principle are uncleared or not discussed. Moreover, the validity
of the quantum $H$-theorem and the second law of thermodynamics, and the
interpretation of the quantum entropy is still in discussion %[19-26].
\cite{bib:silva2010,bib:deroeck2006,bib:grabert1974,bib:acharya2019,%
bib:kastner2017,bib:gring2012,bib:han2015,bib:das2018}.

In order to contribute to the construction of a consistent classical and
quantum $H$-theorem, and a new formalism to describe out of equilibrium
systems, we develop a new theoretical basis to describe them. In particular, we
are interested in including a non-homogeneous distribution function in the
mathematical model, and so including non-homogeneous systems in the $H$
theorem. Besides, we are interested in developing the quantum analogous of
Boltzmann's $H$-theorem considering a non-spatial-homogeneous distribution
function and solve the problem of the correspondence principle mentioned above.

The organization of this article is as follows.
In Section~\ref{sec:classical}, we highlight, to our purposes, the most
fundamental assumptions required to proof the Boltzmann $H$-theorem,
we provide an alternative method to obtain the Maxwell-Boltzmann
distribution using the variational method, and an alternative demonstration of
the $H$-theorem for classical systems. In Section~\ref{sec:quantum}
we review the Tolman proposal for the quantum
version of the $H$-theorem (quantum $H$-theorem)
\commr{and how the Bose-Einstein and Fermi-Dirac distributions are treated within
this framework},
and our alternative proof of the quantum $H$-theorem. In 
Section~\ref{sec:qccorrespondence}
the classical-quantum correspondence between the
quantum $H$-functional and the classical $H$-functional is analyzed.
In Section~\ref{sec:relaxproc} we explore how relaxation processes occur
in a quantum ideal gas; here we provide a time-evolution equation that could be
considered the analogous of the classical Boltzmann transport equation. Finally,
we discuss some key ideas derived from our approach in Section~\ref{sec:disscussion},
and provide some conclusions in Section~\ref{sec:conclusions}.

>>>>

The theoretical basis and the procedures that allow to describe classical equilibrium systems
are well-stablished. These procedures can be applied to treat a wide range of the systems
present in nature, and include both the macroscopic phenomenological methods (thermodynamics)
as well as the microscopic description (statistical mechanics). For instance,
in the Kinetic Theory of gases, the behavior of a dilute classical gas
is described through the Boltzmann transport equation \cite{bib:huang},
and the time-evolution of a system towards the equilibrium is proved by
means of the Boltzmann $H$-theorem, wherein the
spatial-homogeneous distribution function hypothesis plays a crucial role.

In contrast, for quantum equilibrium systems, the construction of a framework,
with the same level of
success and universality as the classical version, is still an open problem. For instance,
in order to obtain a complete correspondence principle, between classical mechanics
and quantum mechanics, the quantum analogous of both the Boltzmann $H$-theorem
and the Boltzmann transport equation have not attained a fully accepted form.
In this context, Tolman was one of the earliest physicists to propose a quantum
$H$-theorem \cite{bib:tolman}, using probability transition relations, the random
phases hypothesis, and a spatially homogeneous distribution function.
Also, Tolman proposed a potential quantum analogous of the transport equation,
in terms of the occupation number,
by applying time perturbation theory. Further attempts, under the quantum operator
formalism, have addressed the description of quantum transport phenomena by means of the Hamiltonian
of the system and the master equation (which here is the analogous
of the Boltzmann transport equation)
\cite{bib:gorban2014,bib:bennaim2017,bib:tolman,bib:li2019}. However, these approaches 
are not consistent with the classical-quantum correspondence principle.
In the same line of thought, some authors have proposed $H$-functionals
and attempted to proof a
quantum $H$-theorem \cite{bib:silva2010,bib:deroeck2006,bib:grabert1974}.
However, whether or not the homogeneous distribution function hypothesis is assumed, or
if their framework obeys the correspondence principle is unclear or not discussed whatsoever.
Furthermore, the general validity
of the quantum $H$-theorem, the second law of thermodynamics, and the
interpretation of the quantum entropy is still in discussion %[19-26].
\cite{bib:silva2010,bib:deroeck2006,bib:grabert1974,bib:acharya2019,%
bib:kastner2017,bib:gring2012,bib:han2015,bib:das2018}.

On the other hand, the framework to describe spatially non-homogeneous systems is still
under construction, although several approaches have been developed. For instance,
the celebrated Onsager formulation (\textit{aka} linear thermodynamics)
\cite{bib:keizer1987,bib:onsager1931} have been successful in describing irreversible chemical 
and physical phenomena. Nonetheless, some aspects such as describing the internal behavior of glasses
\cite{bib:zanotto2018} and the entropy measurement
\cite{bib:schmelzer2018,bib:nemilov2018} cannot be completely addressed with linear thermodynamics.

In addition, some aspects regarding the classical $H$-theorem and the
Boltzmann $H$-functional require to be revised, for the purpose of improving their mutual consistency.
For example, the modification of the $H$-theorem to include phenomena stemming from stochastic
trajectories, violations of the second law of thermodynamics, and the relation
between Shannon's measure of information and the Boltzmann's entropy
\cite{bib:nemilov2018,bib:keizer1987,bib:onsager1931,bib:brown2008,bib:dragoljub2009}.

In order to contribute to the construction of a consistent classical and
quantum $H$-theorem, within a formalism that describes out-of-equilibrium
non-homogeneous systems, in this article we propose a new theoretical framework.
Specifically, for both classical and quantum systems, we include non-homogeneous distribution functions in
the $H$-functionals, and consider non-homogeneous systems in the proofs of the resulting
$H$-theorems. Our proposed $H$-functionals obey
the correspondence principle, but more importantly, these functionals describe
the time-evolution of spatially non-homogeneous
systems towards equilibrium.

The organization of this article is as follows.
In Section~\ref{sec:classical}, we highlight, to our purposes, the most
fundamental assumptions required to proof the Boltzmann $H$-theorem,
we provide an alternative method to obtain the Maxwell-Boltzmann
distribution using the variational method, and the demonstration of
an alternative $H$-theorem for classical systems. In Section~\ref{sec:quantum}
we review the Tolman proposal for the quantum
version of the $H$-theorem (quantum $H$-theorem)
\commr{and how the Bose-Einstein and Fermi-Dirac distributions are treated within
this framework},
and our alternative proof of the quantum $H$-theorem. In 
Section~\ref{sec:qccorrespondence}
the classical-quantum correspondence between the
quantum $H$-functional and the classical $H$-functional is analyzed.
In Section~\ref{sec:relaxproc} we explore how relaxation processes occur
in a quantum ideal gas; here we provide a time-evolution equation that could be
considered the analogous of the classical Boltzmann transport equation. Finally,
we discuss some key ideas derived from our approach and close with a brief summary
in Section~\ref{sec:disscussion}.
